### 2021-07-04 rad-icfp-2021/intro.tex
rewrote contribution

@@

@@

Our work provides a sort of compilation scheme from a source functional language to a target functional language for reverse-mode differentiation.
To emphasize some key ideas and for presentation purposes, we present our work on a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show some useful generalisation in Section~\ref{sec:generalisation}.

The key ideas of our work can be summarised by saying that reverse-mode differentiation
\begin{itemize}
   \item can be made efficient way more easily on unary operators
   \item can be done efficiently and purely using second-order array operators for second-order array operators (e.g. !map,foldl!) 
   \item can be made efficient via standard functional optimisation techniques
\end{itemize}

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation to a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
        % \begin{center}
        %     \begin{tabular}{l}
        %         !let w1 = x1 * x2 in!\\ 
        %         !let w2 = w1 * x1 in!\\
        %         !let y1 = 0 in!\\
        %         !let y2 = 0 in!\\
        %         !let y3 = 0 in!\\
        %         !let y4 = 0 in!\\
        %         !let y5 = 1 in!\\
        %         !let y1'! != y1+w1*y5 in!\\
        %         !let y4'! != y4+x1*y5 in!\\
        %         !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
        %     \end{tabular}
        % \end{center}
    \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

>>>>

TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.

@@

TODO: same as above, maybe rewrite for more theory audience.

@@

We propose a source-code transformation on simple purely functional language for a purely functional reverse-mode differentiation.
Our transformation is decomposed in a sort of compilation scheme through a novel intermediate representation, unary form (UNF).
To emphasize some key ideas and for presentation purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.

We summarize the key ideas of our work. Reverse-mode can be made efficient
\begin{itemize}
   \item easily if there are only unary operators
   \item in a purely functional way
   \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
   \item via standard functional optimization techniques
\end{itemize}

Our contribution makes these claims precise. In detail we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\ref{sub:sourcelang})
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\ref{sec:generalisation})
    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\ref{sec:unf})
    \item a transformation to and from the I.R. from the main language \ref{sub:transformations to and from UNF}
    \item a simple reverse-mode transformation on the I.R. (\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\ref{sub:Macro for pure reverse mode transformation})
    \item denotational semantics of our languages and transformations using multicategories and concategories \ref{sec:correctness}
    \item correctness of the transformations \ref{sec:correctness}
    \item complexity guarantees \ref{sec:complexity}
    \item how efficiency is recovered via standard optimization techniques \ref{sub:Optimisations}
    \item several additions to the source language (\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

After some general partial evaluation techniques that will be detailed further in the paper, we obtain

    \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}
